---
description: "Base R vs ggplot2 in 2026: measured draw speed, session startup cost, CRAN dependency counts and container sizes, with a decision tree for choosing between them."
---

# Why Base R Graphics in 2026? {#why-base}

## Introduction

Base graphics have a reputation problem. The usual first question about them is
some version of *isn't this the old way?*, and the usual answer is a shrug. This
chapter argues the opposite: in 2026, base graphics are the correct first tool
for a specific and fairly large set of jobs, and this book is that tool's
documentation.

In this chapter, we will

- establish what base graphics actually is, and what ships with R
- measure how fast they draw, and how that measurement is usually faked
- measure what a plotting session costs to start
- count the dependencies that `ggplot2` adds and base graphics does not
- walk a decision tree for choosing between the two

We will do this with numbers rather than adjectives. Every figure in this
chapter is produced by a script in `scripts/`, stored as a CSV in `data/`, and
committed. Nothing here runs when the book is built, which is deliberate: a
book that argues for zero dependencies cannot quietly require two of them to
prove the point. If a number in this chapter is ever wrong, you can re-run the
script that produced it and find out which of us was wrong.

## What "base graphics" means

Base graphics is the set of plotting functions that ship inside R itself, in
the `graphics` package. There is nothing to install.

```{r whatis, eval=FALSE}
?graphics
```

The package has been part of R since before R 1.0 and its core has barely
changed, which is either its greatest strength or its most damning indictment
depending on who is talking. The reason it survives is that it does one thing
very well: it takes coordinates and draws them. Everything else in this book is
an elaboration on that idea.

The key practical fact is this one:

```{r base-ready}
plot(mtcars$disp, mtcars$mpg)
```

No `library()` call. `graphics` is one of the packages R attaches at startup,
alongside `stats`, `grDevices`, `utils` and `methods`. It is not merely
available, it is already loaded, which is a stronger guarantee than "cheap to
install". We will quantify that in a moment.

The three systems you are likely to meet in R are:

- **base graphics** (`graphics`) — in R, always present, coordinates in, marks
  out
- **`ggplot2`** — a large, opinionated package built on the grammar of graphics,
  with a large dependency tree
- **`lattice`** — a Trellis-style system for multivariate data, part of R's
  recommended packages

This book is about the first one. Chapter 2 surveys all three.

### Why the naming matters

Base graphics is frequently described as "low-level". That is accurate and
slightly misleading in equal measure. The *drawing* primitives are low-level:
`plot()`, `points()`, `lines()`, `text()`, `polygon()`, `arrows()` and `rect()`
put marks at coordinates you choose. But the high-level functions you actually
call most often — `barplot()`, `boxplot()`, `hist()`, `pairs()`, `plot()` as a
scatterplot, `dotchart()`, `stripchart()`, `mosaicplot()`, `image()` — are
convenience wrappers that pick sensible defaults and let you override any of
them.

So base graphics is low-level in the sense that **you stay in control**. Nothing
is computed behind your back, no scale is transformed without being named, no
grouping is inferred. That control is the actual product here, and it is what
the rest of the book teaches you to spend.

## Speed

The claim to test is that base graphics are fast, and the interesting part is
that the claim is usually tested wrong.

We drew 10,000 and 100,000 points three ways, timing a real render each time:

- **`base`** — `plot(x, y)`
- **`ggplot2_construct`** — `ggplot(df, aes(x, y))`, with no `print()`
- **`ggplot2_render`** — the same, plus `print()` so it actually draws

```{r bench}
bench <- read.csv('data/bench-base-vs-ggplot2.csv')
bench
```

Three approaches, two sizes, median of 15 runs. Now the part that matters:

> The middle row measures almost nothing. `ggplot()` *builds an object*. It does
> not draw. Until something calls `print()` on it, no marks are laid down and no
> pixels or vector output are produced. So `ggplot2_construct` is timing R's
> object system, not its graphics engine. If you have ever read that "ggplot2 is
> faster than base R", that is very likely the number you read.

Both sides here genuinely draw. Base renders to `pdf(NULL)`, a device that
accepts and discards output, so the plotting work happens with no window and no
file.

```{r bench-plot, fig.align='center', fig.width=9, fig.height=4.5}
#| label: fig-bench
#| fig-cap: "Median time to draw, by engine and problem size"
init <- par(no.readonly = TRUE)

sizes <- sort(unique(bench$n))
approaches <- c('base', 'ggplot2_construct', 'ggplot2_render')
cols <- c(base = '#0072B2', ggplot2_construct = '#999999',
          ggplot2_render = '#D55E00')
labs <- c(base = 'base plot()',
          ggplot2_construct = 'ggplot2\nconstruct only',
          ggplot2_render = 'ggplot2\nprint()')

# 1e+05 formats as "1e+05" unless scientific is switched off, which is not a
# caption anyone wants on a figure.
fmt_n <- function(x) format(x, big.mark = ',', scientific = FALSE, trim = TRUE)

draw_panel <- function(n) {
  v <- bench$median_ms[match(approaches, bench$approach[bench$n == n])]
  # barplot stacks bottom-up, so rev() puts base first and reads top-down in the
  # same order as the prose does. Titles are kept short because a long one
  # overflows the panel once the left margin has to clear two-line labels.
  barplot(rev(v), horiz = TRUE, log = 'x', names.arg = rev(labs), las = 1,
          cex.names = 0.8, col = rev(cols[approaches]), border = NA, space = 0.4,
          xlim = c(1, 3000), main = sprintf('%s points', fmt_n(n)),
          xlab = 'median time to draw, milliseconds (log scale)')
}

# Left margin has to clear two-line category labels plus the axis; the default
# silently clipped them.
par(mfrow = c(1, 2), mar = c(4.5, 6, 3, 1))
draw_panel(sizes[1])
draw_panel(sizes[2])

par(init)
```

@fig-bench shows the result. At 10,000 points, constructing the ggplot2 object is
about as fast as drawing the whole thing in base graphics — and drawing it is
roughly **80 times slower than drawing it in base**. At 100,000 points the gap is
worse, and base graphics has not become slower so much as ggplot2 has failed to
scale the way the underlying drawing work does.

Memory tells a similar story:

```{r bench-mem, fig.align='center', fig.width=6.5, fig.height=4.5}
#| label: fig-bench-mem
#| fig-cap: "Peak memory to draw 100,000 points, by engine"
init <- par(no.readonly = TRUE)
n <- max(bench$n)
v <- bench$peak_mem_mb[match(approaches, bench$approach[bench$n == n])]
barplot(rev(v), horiz = TRUE, names.arg = rev(labs), las = 1, cex.names = 0.8,
        col = rev(cols[approaches]), border = NA, space = 0.4,
        main = sprintf('Peak memory to draw (%s points)', fmt_n(n)),
        xlab = 'MB')
par(init)
```

Base peaks at 31 MB where ggplot2 peaks at 77 MB. Note the middle bar again: the
object-only path uses *less* memory than base, which is a direct consequence of
not drawing anything.

> These are absolute timings from one machine — Windows, R 4.5.2, `x86_64-w64-mingw32`,
> ggplot2 4.0.1 — and the script is committed at `scripts/bench.R`, so the numbers
> are reproducible rather than remembered. Treat them as an order of magnitude.
> What travels across machines is the *shape* of the result: one engine does the
> drawing work immediately, and the other builds a description of a plot that
> something else has to be asked to render. That extra step is the cost, and it
> is not a constant — it grows with what you ask for.

That last clause is the real argument. Base graphics is a direct-mode engine:
your call draws. ggplot2 is deferred: your call records intent, and rendering is
a separate step performed later, on a plot that may have grown additional layers
in the meantime. Deferred is the right design for iteration and exploration. It
is the wrong design when the answer is a fixed number in a fixed report.

## The cost of a session

Speed per plot only matters if you pay to start plotting at all. Here is what it
costs to bring a session up, measured as wall clock around a whole `Rscript`
process.

```{r startup}
startup <- read.csv('data/startup-latency.csv')
startup[, c('approach', 'wall_ms', 'spread_ms', 'overhead_ms', 'load_ms', 'namespaces')]
```

Three rows, and they are not close.

```{r startup-plot, fig.align='center', fig.width=7, fig.height=4}
#| label: fig-startup
#| fig-cap: "Wall clock to start a plotting session, by engine"
init <- par(no.readonly = TRUE)
op <- par(mar = c(4.5, 5, 3, 1))
barplot(
  startup$wall_ms,
  horiz = TRUE, names.arg = startup$approach, las = 1, cex.names = 0.9,
  col = c('#999999', '#0072B2', '#D55E00'), border = NA, space = 0.4,
  xlim = c(0, max(startup$wall_ms) * 1.15),
  main = 'Wall clock to start a plotting session (ms)',
  xlab = 'milliseconds, median of 15 runs'
)
par(op)
par(init)
```

`library(graphics)` costs nothing measurable, and the reason is visible in the
last column: the namespace count does not move from 8 to 8. `graphics` is already
attached at startup, so the call is a no-op. That is not "cheap" in the way a
cached download is cheap — it is *free*, structurally.

`library(ggplot2)` takes about six seconds on this machine and pulls the loaded
namespace count from 8 to 29. Twenty-one packages.

> The `overhead_ms` column for `graphics` is near zero and can land slightly
> negative; `spread_ms` is the run-to-run range, and it is larger than the
> difference. The honest reading is "below the noise floor", not "measurably
> fast". Again, one machine — but the namespace column is not machine-specific,
> and that is the number to remember.

## The dependency count

The strongest form of the zero-dependency claim is not an adjective but a count.
Base graphics needs zero packages that are not already in R. Packages that build
on it inherit that property, because there is nothing to pull in.

We read this from CRAN's own metadata, counting each package's declared
`Depends`, `Imports` and `LinkingTo` entries and discarding everything that ships
with R:

```{r cran}
deps <- read.csv('data/cran-deps.csv')
deps[, c('package', 'role', 'version', 'n_nonbase', 'imports_ggplot2')]
```

Sort that by dependency count and a pattern appears (@fig-cran):

```{r cran-plot, fig.align='center', fig.width=7.5, fig.height=5}
#| label: fig-cran
#| fig-cap: "Declared non-base CRAN dependencies by package"
init <- par(no.readonly = TRUE)

d <- deps[order(deps$n_nonbase, deps$package), ]
lab <- sprintf('%-16s %2d', d$package, d$n_nonbase)
# Axis runs to the observed maximum. An earlier version capped it at 12 and
# warned about capping, which is a confusing thing to say when the cap happens
# to equal the largest bar and nothing is actually cut.
op <- par(mar = c(5.5, 9, 3, 1))
bp <- barplot(
  d$n_nonbase, horiz = TRUE, names.arg = FALSE, border = NA,
  space = 0.35, col = ifelse(d$imports_ggplot2, '#D55E00', '#0072B2')
)
axis(2, at = bp, labels = lab, las = 1, tick = FALSE, cex.axis = 0.85)
title(main = 'Non-base CRAN dependencies by package',
      xlab = 'declared Depends / Imports / LinkingTo outside base + recommended')
mtext('orange = also depends on ggplot2', side = 1, line = 3.6,
      cex = 0.75, col = 'grey30')
par(op)
par(init)
```

Read the blue bars. `TeachingDemos`, `plotrix` and base R itself declare **zero**
non-base dependencies, and they still produce publication figures — `plotrix` is
a dedicated graphics add-on. `faraway` needs one. `psych`, which has been drawing
its own plots with base graphics for a decade, needs two.

That is the practical shape of the argument. A base-graphics package on CRAN is
cheap to install, cheap to check, cheap to run in a locked-down environment, and
cheap to keep building in ten years, because its dependency graph is nearly
empty.

And here is the part that would be dishonest to leave out:

> The orange bars include this author's own CRAN packages — `olsrr`, `rfm`, `blorr`
> and `descriptr` — and they all depend on `ggplot2`. They do not use base
> graphics. We checked their sources rather than inferring it from metadata: as of
> 2026-10-04, `rfm/R/rfm-plots.R` and `descriptr/R/ds-plots.R` are entirely
> ggplot2, and so are the diagnostic plots in `olsrr`. Metadata cannot tell you
> what a package draws with; you have to read the code.

So what does that say about base graphics? Read it as a trajectory rather than an
endorsement. Those packages got onto CRAN drawing base graphics, then outgrew it,
exactly as this book predicts they would. The base layer was the on-ramp; ggplot2
is where they landed once the plotting became the product rather than a
by-product. Chapter 2 covers the systems side by side, and Appendix A maps every
argument in this chapter onto its ggplot2 equivalent so you can translate in
either direction.

## What it weighs in a container

Dependency counts are an abstraction. If you ship R in a container — and if you
ship R anywhere locked down, you probably do — the question stops being "how
many packages" and becomes "how many bytes".

Two images from the same publisher make that measurable, because
`rocker/tidyverse` is built *on top of* `rocker/r-ver`: the same R, the same base
system, plus the package layer. The difference between them is therefore the
cost of the tidyverse layer and nothing else, which is a cleaner comparison than
diffing two unrelated images.

```{r docker}
docker <- read.csv('data/docker-sizes.csv')
docker[, c('image', 'tag', 'layer', 'size_mb', 'extra_layer_mb')]
```

```{r docker-plot, fig.align='center', fig.width=7.5, fig.height=4}
init <- par(no.readonly = TRUE)
d <- docker[docker$tag == '4.5.2', ]
# Bottom margin has to clear the x label AND the delta caption below it.
par(mar = c(5.5, 9, 3, 1))
bp <- barplot(d$size_mb, horiz = TRUE, names.arg = FALSE, border = NA,
              space = 0.7, col = c('#0072B2', '#D55E00'))
axis(2, at = bp, labels = d$image, las = 1, tick = FALSE, cex.axis = 0.9)
# Title kept short: a long one overflows the panel once the left margin has to
# clear these labels.
title(main = 'Container image size, R 4.5.2', xlab = 'MB, compressed download')
delta <- d$size_mb[2] - d$size_mb[1]
mtext(sprintf('the tidyverse layer adds %s MB on top of R, a %.2fx increase',
              format(delta, big.mark = ',', scientific = FALSE),
              d$size_mb[2] / d$size_mb[1]),
      side = 1, line = 4.2, cex = 0.75, col = 'grey30')
par(init)
```

A base-R image is 349 MB; adding the tidyverse layer takes it to 924 MB. That
574 MB is not a mistake or a packaging artefact — it is the price of the
dependency tree, paid on every pull, every CI run and every cold start, whether
or not the code being run touches a single one of those packages.

> Two caveats, both recorded in the CSV. These are Docker Hub's reported sizes
> for the whole tag, which is the sum of its **compressed** layers — a real
> number a real user downloads, but not the unpacked size on disk, and the ratio
> is only as fair as the fact that both tags were built at the same time by the
> same publisher. And they are published metadata read over HTTP, not a local
> build: no daemon was involved, which is why this is reproducible on any
> machine, and also why it cannot control the comparison.

Read next to the CRAN table above, the pattern is consistent. Base graphics is
not lighter by a little; it is lighter by an order of magnitude, because it
inherits the empty part of R's dependency graph rather than extending it.

## What base graphics costs

An honest chapter has to state the other side.

- **More typing.** Base graphics is verbose by design. The same plot is longer in
  base than in ggplot2, and it stays that way.
- **No automatic grouping or faceting.** You build panels yourself with
  `par()` or `layout()` and a loop, where ggplot2 has `facet_wrap()`. Chapter 11
  covers this and it is the single biggest readability jump in the book.
- **No automatic scale transformation.** Nothing is transformed unless you name
  it. This is a feature when you want control and a cost when you do not.
- **A smaller ecosystem.** The reference grid, the grammar extensions, the
  themes, the colour scales: all of that ecosystem is ggplot2's, not base's.
- **Older idiom.** Base graphics code from 2001 still runs. That cuts both ways.

## A decision tree

Most of the above reduces to one question: who decides what the plot looks like?

```{r decision, fig.align='center', fig.width=8, fig.height=5.5}
#| label: fig-decision
#| fig-cap: "Choosing between base graphics and ggplot2"
init <- par(no.readonly = TRUE)
# A flowchart is a picture, not a data plot: no axes, and the coordinates are
# laid out by hand. ps is set explicitly because text size otherwise depends on
# the device knitr happened to open, which is how a hand-placed diagram ends up
# with labels larger than its boxes.
par(xpd = NA, mar = c(0, 0, 0, 0), ps = 9)
plot.new()
plot.window(xlim = c(0, 100), ylim = c(12, 104))

box <- function(x, y, w, h, label, sub, fill) {
  rect(x, y - h / 2, x + w, y + h / 2, col = fill, border = 'grey25', lwd = 1.5)
  text(x + w / 2, y + (if (nzchar(sub)) 3.2 else 0), label, font = 2, cex = 0.92)
  if (nzchar(sub)) {
    text(x + w / 2, y - 4.6, sub, cex = 0.66, col = 'grey20')
  }
}
yes_no <- function(x, y, lab) text(x, y, lab, cex = 0.7, col = 'grey25')

Q <- '#DCE9F5'   # question
G <- '#F7E2D5'   # ggplot2 outcome
B <- '#DFF0E3'   # base outcome

box(4, 90, 52, 15, 'Is the plot final?',
    'One answer, written once, no iteration', Q)
box(4, 57, 52, 19, 'Does it have to run somewhere\nyou do not control?',
    'Locked-down server, CRAN check, minimal container', Q)
box(64, 90, 34, 15, 'ggplot2', 'Explore, iterate, facet, publish', G)
box(64, 57, 34, 15, 'ggplot2', 'Communicate, share, present', G)
box(22, 22, 52, 17, 'Base graphics',
    'Package internals, batch reports,\nexact control over the marks', B)

arrows(56, 90, 63.5, 90, length = 0.07, lwd = 1.6)
arrows(30, 82.4, 30, 66.7, length = 0.07, lwd = 1.6)
arrows(56, 57, 63.5, 57, length = 0.07, lwd = 1.6)
arrows(30, 47.4, 40, 30.7, length = 0.07, lwd = 1.6)

yes_no(59.7, 93, 'yes')
yes_no(32.4, 74.5, 'no')
yes_no(59.7, 60, 'yes')
yes_no(24, 39, 'no')

par(init)
```

Two questions, and note that "final" is doing more work than "simple". A plot
you will iterate on twenty times belongs to ggplot2 even if it is a single
scatterplot, because the cost of base graphics is paid on every edit. A plot you
will write once and run ten thousand times belongs to base graphics even if it
is complicated, because that is where the per-plot cost actually lands.

Note also that the second question overrides the first. A plot that must run on a
server you do not control is a base-graphics plot even if you would rather be
iterating, because "install ggplot2 on this locked-down box" is not a decision
you get to make.

That tree is a starting point, not a rulebook. Plenty of working code mixes the
two: base graphics for the bulk of a report, one ggplot2 plot where the faceting
carries the argument. The point is that the choice should be made deliberately,
per plot, rather than by default.

## When you have read this chapter

You now have four measured numbers and a decision tree. In the rest of the book
we stop arguing and start drawing, and every chapter is written to the assumption
that you want the marks exactly where you put them.

Chapter 2 surveys the three R graphics systems and walks the six dispatch cases
of `plot()`. From Chapter 3 onward the argument is one argument at a time: here
is the bare call, here is what changing `pch` does, here is everything combined.
Nothing in this book requires installing anything.